
```{r,results=FALSE,echo=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
source("common.R")
library(Hmisc) # binconf
(load(file=file_rdata))

################################################################################

## from https://www.tandfonline.com/doi/abs/10.1080/00031305.2018.1473796?af=R&journalCode=utas20
## functions for upper and lower limits of MI-Wilson interval
#Upper limit function for the MI-Wilson interval for multiple imputation
upper.limit= function(z, n, r.m, mean_q) {
  ul = ((((2*mean_q) + ((z^2)/n) + (((z^2) * r.m)/n))/
           (2*(1 + ((z^2)/n) + (((z^2)*r.m)/n)))) +
          sqrt(((((2*mean_q) + ((z^2)/n) + (((z^2)*r.m)/n))^2)/
                  (4*(1 + ((z^2)/n) + (((z^2)*r.m)/n))^2)) -
                 ((mean_q^2)/(1 + ((z^2)/n) + (((z^2)*r.m)/n)))))
  return(ul)
}

#Lower limit function for the MI-Wilson interval for multiple imputation
lower.limit= function(z, n, r.m, mean_q) {
  ll = ((((2*mean_q) + ((z^2)/n) + (((z^2) * r.m)/n))/
           (2*(1 + ((z^2)/n) + (((z^2)*r.m)/n)))) -
          sqrt(((((2*mean_q) + ((z^2)/n) + (((z^2)*r.m)/n))^2)/
                  (4*(1 + ((z^2)/n) + (((z^2)*r.m)/n))^2)) -
                 ((mean_q^2)/(1 + ((z^2)/n) + (((z^2)*r.m)/n)))))
  return(ll)
}

x <- file.path(d,"predictions_on_test_set_no_12.rds")  %>% readRDS()
with(x,table(group,type))
x[,type:=sub("\\."," ",type)]
x[,Sample.ID:=gsub("wk|WK","Wk",Sample.ID)] # typo
sample(x$Sample.ID,100)
x[grep("Wk",Sample.ID), Week:=sub(".*(Wk[0-9]+).*","\\1",Sample.ID) ]
x[,prob.ENS:=(prob.SVM+prob.LDA)/2]
table(x$Week)
x=x[Week!="Wk14"]

## add 3sd double positives
spike.3sd <- with(m[type=="Historical controls"], mean(SPIKE) + 3*sd(SPIKE))
rbd.3sd <- with(m[type=="Historical controls"], mean(RBD) + 3*sd(RBD))
x[,prob.3sd:=as.numeric(SPIKE > spike.3sd & RBD > rbd.3sd)]
m[,prob.3sd:=as.numeric(SPIKE > spike.3sd & RBD > rbd.3sd)]

mx <- m[type %in% c("Historical controls","COVID")]
library(viridis)
```

We used the ensemble SVM-LDA learner to estimate the proportion of 1100 pregnant volunteers and blood donors sampled from weeks 17-21 of 2020 showing seropositivity to S and RBD.

```{r}
with(x, table(Week,type))
```

This plot shows *a* the training data (PCR+ COVID patients and historical controls sampled in 2019) *b* the test data (pregnant volunteers and blood donors) and *c* an overlay of the two.  Colour scale red (low) to blue (high) indicates the estimated probability of seropositivity.

```{r}
p3 <- ggplot(mapping=aes(y=log2(SPIKE),x=log2(RBD))) +
  geom_point(aes(pch=type),data=mx,col="grey") +
  geom_point(aes(col=prob.ENS),data=x,pch=17) +
  theme(legend.position="bottom") +
  ## scale_colour_viridis() +
  scale_colour_gradient2(midpoint=0.5,mid="orange") + 
  scale_shape_manual(values=c("Historical controls"=21,"COVID"=20))
p2 <- ggplot(mapping=aes(y=log2(SPIKE),x=log2(RBD))) +
  ## geom_point(aes(pch=type),data=mx,col="grey") +
  geom_point(aes(col=prob.ENS),data=x,pch=17) +
  theme(legend.position="bottom") +
  ## scale_colour_viridis() +
  scale_colour_gradient2(midpoint=0.5,mid="orange") + 
  scale_shape_manual(values=c("Historical controls"=21,"COVID"=20))
p1 <- ggplot(mapping=aes(y=log2(SPIKE),x=log2(RBD))) +
  geom_point(aes(pch=type),data=mx,col="grey") +
  theme(legend.position="bottom") +
  ## scale_colour_viridis() +
  scale_colour_gradient2(midpoint=0.5,mid="orange",
    limits=c(0,1),breaks=c(0,1),labels=c(0,1),
                         guide=guide_colorbar(labels=NULL,draw.ulim=TRUE,draw.llim=TRUE)) +
  ## geom_point(aes(col=prob.ENS),data=x,pch=4) +
  scale_shape_manual(values=c("Historical controls"=21,"COVID"=20))
ens <- plot_grid(p1,p2,p3,nrow=1,align="hv",labels=c("a","b","c"))
ens
## p3 <- ggplot(mapping=aes(y=log2(SPIKE),x=log2(RBD))) +
##   geom_point(aes(pch=type),data=mx,col="grey") +
##   geom_point(aes(col=prob.3sd),data=x,pch=17) +
##   theme(legend.position="bottom") +
##   ## scale_colour_viridis() +
##   scale_colour_gradient2(midpoint=0.5,mid="orange") + 
##   scale_shape_manual(values=c("Historical controls"=21,"COVID"=20)) +
## geom_vline(xintercept=log2(rbd.3sd)) +
##   geom_hline(yintercept=log2(spike.3sd))
## p2 <- ggplot(mapping=aes(y=log2(SPIKE),x=log2(RBD))) +
##   ## geom_point(aes(pch=type),data=mx,col="grey") +
##   geom_point(aes(col=prob.3sd),data=x,pch=17) +
##   theme(legend.position="bottom") +
##   ## scale_colour_viridis() +
##   scale_colour_gradient2(midpoint=0.5,mid="orange") + 
##   scale_shape_manual(values=c("Historical controls"=21,"COVID"=20)) +
##   geom_vline(xintercept=log2(rbd.3sd)) +
##   geom_hline(yintercept=log2(spike.3sd))
## p1 <- ggplot(mapping=aes(y=log2(SPIKE),x=log2(RBD))) +
##   geom_point(aes(pch=type),data=mx,col="grey") +
##   theme(legend.position="bottom") +
##   ## scale_colour_viridis() +
##   scale_colour_gradient2(midpoint=0.5,mid="orange") + 
##   ## geom_point(aes(col=prob.LOG),data=x,pch=4) +
##   scale_shape_manual(values=c("Historical controls"=21,"COVID"=20))
## sd3 <- plot_grid(p1,p2,p3,nrow=1)
## sd3

```

```{r,echo=FALSE,results=FALSE}
## estimate proportion + CI

fitone <- function(x,probvar="prob.log",byvars=c("type","Week"),M=1000) {
  set.seed(42) # reprocibility
  summ <- lapply(1:M, function(j) {
    x[,y:=runif(.N) <= x[[probvar]] ] # simulate case status conditional on fit
    x[,.(yhat=sum(y),phat=mean(y),varp=mean(y)*mean(1-y)/.N,N=.N,j=j),by=byvars]
  })  %>% rbindlist()
  ## wilson's interval with observed data for comparison
  wilsons <- with(summ, binconf(yhat,N))
  summ <- cbind(summ,wilsons)
  summ2 <- summ[,.(N=mean(N),
                   qm=mean(phat),um=mean(varp),bm=var(phat), # MI
                   Est=mean(PointEst),Lower=mean(Lower),Upper=mean(Upper)), # observed data
                by=byvars]
  summ2[,Tm:=(1+1/M)*bm + um]
  if(any(summ2$bm==0)) { # all realisations the same - doesn't happen, just in case
    summ2[bm==0, c("v","rm"):=list(1000000000,0)]
  }
  summ2[bm!=0,
        c("v","rm"):=list((M-1)*(1+(um/((1+ (1/M))*bm)))^2, (1 + (1/M))*(bm/um))]
  summ2[,ll:=lower.limit(z = qt(0.975, df = v), n = N, r.m = rm,
                             mean_q = qm),by=byvars]
  summ2[,ul:=upper.limit(z = qt(0.975, df = v), n = N, r.m = rm,
                             mean_q = qm),by=byvars]
  options(digits=2)
  summ2[,c(byvars,c("N","Est","Lower","Upper","qm","ll","ul")),with=FALSE]
}

## per group and week
m <- "ENS"
probs1 <- fitone(x,paste0("prob.",m),byvars=c("type","Week"))
probs1[order(type,Week),.(type,Week,N,estimate=qm,lower.ci=ll,upper.ci=ul)]

## per week
probs2 <- fitone(x,paste0("prob.",m),byvars="Week")[,type:="Combined"]
probs2[order(Week),.(Week,N,estimate=qm,lower.ci=ll,upper.ci=ul)]

## per group
probs3 <- fitone(x,paste0("prob.",m),byvars="type")[,Week:="Combined"]
probs3[order(type),.(type,N,estimate=qm,lower.ci=ll,upper.ci=ul)]

## overall
x[,overall:=TRUE]
probs4 <- fitone(x, paste0("prob.",m),byvars="overall")[,Week:="Combined"][,type:="Combined"]

## combine
probs <- rbind(probs1,probs2,probs3,probs4,fill=TRUE)[,method:=m]
probs[,week_num:=as.numeric(sub("Wk","",Week))]
probs[,type:=factor(type,levels=c("Blood donors","Pregnant volunteers","Combined"))]

pred <- probs
with(pred,table(Week,type))

tmp <- dcast(pred[method=="ENS" & Week!="Wk14"],
      Week ~ type, value.var="Est")
tmp
setnames(tmp,make.names(names(tmp)))
tmp[,.(bd=mean(Blood.donors),pv=mean(Pregnant.volunteers))]

pred[,week_num2:=ifelse(week_num < 14, week_num + 52, week_num)]

ggplot(pred[type=="Combined" & Week!="Combined"],aes(x=week_num2,y=qm,ymin=ll,ymax=ul#,col=type
                                  )) +
  ## geom_smooth(se=FALSE,
  ##             ## method="lm",formula=y~poly(x,3),
  ##                                       span=0.8,
  ##   alpha=0.2,
  ##             linetype="dotted") +
  geom_pointrange() +
  ## facet_wrap(~type) +
  facet_grid(type ~ method) +
  background_grid(major="y")
```

We estimated the proportion of positives in each group using multiple imputation, using the  MI-Wilson method to estimate the confidence interval. Results were similar between groups.

```{r}
library(seaborn)
ggplot(pred[type!="Combined" & Week!="Combined"],aes(x=week_num2+as.numeric(factor(type))/10,
                                  y=qm,ymin=ll,ymax=ul,col=type)) +
  geom_pointrange() +
  labs(x="Week",y="Estimated fraction positive") +
  background_grid(major="y") + scale_colour_seaborn()

ggplot(pred[type=="Combined" & Week!="Combined"],aes(x=week_num2,
                                  y=qm,ymin=ll,ymax=ul)) +
  geom_pointrange() +
  labs(x="Week",y="Estimated fraction positive") +
  background_grid(major="y") + scale_colour_seaborn()

options(digits=6)
kable(probs[][order(type,Week),.(type,Week,N,estimate=qm,lower.ci=ll,upper.ci=ul)])
fwrite(probs[][order(type,Week),.(type,Week,N,estimate=qm,lower.ci=ll,upper.ci=ul)],
       file=file.path(d,"estimates.csv"))
fwrite(probs[][order(type,Week),.(type,Week,N,estimate=qm,lower.ci=ll,upper.ci=ul)],
       file=file.path("estimates.csv"))
```
